Journey: A Study of Scalabilty of Memory Management in Linux (TODO: should we get a new title?)

### 2017-04-27 paper.tex
Some changes to paper

\title{Journey: A Study in Unhelpful Titles (TODO: should we get a new title?)}
\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada\\\texttt{\{markm, hp, spai\}@cs.wisc.edu}}
\date{}

@@

The Linux kernel keeps track of which portions of a process's address space
have been allocated using the \texttt{vm\_area\_struct}. Because processes may
map a number of regions, including shared objects, stack, heap, text, and
anonymous regions, the scalability of this structure is critical to Linux's
performance.

When a process maps a region, a \texttt{vm\_area\_struct} is created, but no
memory is allocated. When a process first touches a page, a page fault occurs
and the processor traps into a page fault handler, which can allocate memory
and set up a page table entry. This is called demand paging.

For demand paging to work efficiently, the correct \texttt{vm\_area\_struct}
needs to be efficient to find. Linux caches the most recently used region and
organizes all regions into a balanced binary tree for fast lookup.  Moreover,
adjacent regions with similar permissions and properties are merged in an
effort to keep bookkeeping costs low.

(TODO: explain struct page?)

We measure the memory overhead and latency of the Linux Memory Management
system's use of \texttt{vm\_area\_struct}, which represents a region of a
process's virtual address space.  
We do not measure the overhead of \texttt{struct page} because it is known that
there is one struct for each physical page. We do not measure the overhead of
page tables because this can be calculated from the size of the virtual memory
space used by the process.  When a program makes a memory management system
calls such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap},
\texttt{mprotect}, \texttt{mlock}, or \texttt{madvise}, the kernel is creating,
changing, or removing \texttt{vm\_area\_struct}s; thus, measuring the
scalability of Linux memory management requires measuring the overhead of these
structs.

@@

First, when a continuous \texttt{mmap} (with the same permissions) happens the
kernel memory allocator does not need to create a new \texttt{vm\_area\_struct},
instead it modifies the adjacent \texttt{vm\_area\_struct}s to account for the
newly allocated space. This obviates the need to use the slab allocator to create
a new struct.

Second, for continuous \texttt{mmap} operations, the time taken is always the
lowest compared to other benchmarks.  This is because the kernel caches the most
recently used \texttt{vm\_area\_struct}, avoiding a red-black tree search. In
the continuous benchmark, this cached struct is always conveniently used.

@@

The frag benchmark takes the most time of any benchmark. When a region of the
address space is being expanded, the kernel is forced to unmap it and map a new
area somewhere else.  This operation needs creation of a new memory area struct.
The kernel also must look through the red-black tree, though it never finds a
node which can be expanded to include the new mapping. Then, an insertion on the
tree is needed, possibly including rebalancing. Moreover, the exisiting texttt{
vm\_area\_struct} might need to be deleted and removed from the tree. All of these
operations ensure that frag mmap pattern takes the most amount of time.

The strided workload is very similar to frag because it will also force the
kerenl to always create a new mapping and insert it into the tree after
traversing the tree first. It, however, does not need to delete an existing
mapping so it is faster than the frag benchmark.

Finally, rand will also search through the red-black tree first but there is a
reasonably high probability that it will find an existing mapping adjacent to
the desired mapping which can be expanded, obviating the need to create a new
struct and insert it into red-black tree. Therefore, most of the times the time 
taken by rand benchmark is lower than strided.

@@

The overall behavior is as expected. In the random benchmark, as before, all
memory is exhausted quickly, and the benchmark induces swapping soon after the
benchmark begins. Many memory accesses thus take milliseconds to complete
because they require paging in or out, which in turn depends on disk latency.
Even so, the majority of benchmark run time comes from a minority of accesses.
Figure \ref{fig:swap_time_rand_cdf} shows that as much as 95\% of memory
accesses occur on the micro-second scale. Of the remaining 5\%, almost all
accesses complete in less than 20ms. Moreover, notice that the size of the
swapfile makes very little difference in this workload because the disk has to
do random seeks regardless of swapfile size.

@@

We also find that in the continuous benchmark, a larger swapfile actually has
slight more of these slow memory accesses. We believe this has to do with the
fact that the large swapfile has less disk locality, and so causes more disk
seeks when paging out. Figure \ref{fig:swap_time_cdf} shows that with a larger
swapfile, the continuous benchmark has a significantly higher number of
high-latency accesses as compared with the smaller swapfile.

@@

Large swapfiles should be created with disk locality or on SSDs.

Looking at 99.99-th percentile latencies is important because they can have a
lot of impact on overall run time.

>>>>

\title{Journey: A Study of Scalabilty of Memory Management in Linux (TODO: should we get a new title?)}
\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada\\\texttt{\{markm, hp, spai\}@cs.wisc.edu}}
\date{}

@@

The Linux kernel keeps track of which portions of a process's address space
have been allocated using the \texttt{vm\_area\_struct} data structure. Because
processes may map a number of regions, including shared objects, stack, heap,
text, and anonymous regions, the scalability of this structure is critical to
Linux's performance.

When a process maps a region, a \texttt{vm\_area\_struct} is created, but no
memory is allocated. When a process first touches a page, a page fault occurs
and the processor traps to a page fault handler, which allocates memory and
sets up a page table entry for a valid access. This is called demand paging.

For demand paging to work efficiently, the correct \texttt{vm\_area\_struct}
needs to be efficient to find. Linux caches the most recently used region and
organizes all regions into a red-black tree for fast lookup.  Moreover,
adjacent regions with similar permissions and properties are merged in an
effort to keep bookkeeping costs low.

We measure the memory overhead and latency of the Linux Memory Management
system's use of \texttt{vm\_area\_struct}, which represents a region of a
process's virtual address space. All the physical pages in the system are
managed using the \texttt{struct page} descriptors in the kernel.  We do not
measure the overhead of \texttt{struct page} because it is known that there is
one struct for each physical page. We do not measure the overhead of page
tables because this can be calculated from the size of the virtual memory space
used by the process.  When a program makes a memory management system calls
such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap}, \texttt{mprotect},
\texttt{mlock}, or \texttt{madvise}, the kernel is creating, changing, or
removing \texttt{vm\_area\_struct}s; thus, measuring the scalability of Linux
memory management requires measuring the overhead of these structs.

@@

First, when a continuous \texttt{mmap} (with the same permissions) happens the
kernel memory allocator does not need to create a new \texttt{vm\_area\_struct}.
Instead, it uses the existing \texttt{vm\_area\_struct} to account for the
newly allocated space. This obviates the need to use the slab allocator to create
a new struct.

Second, for continuous \texttt{mmap} operations, the time taken is always the
lowest compared to other benchmarks.  This is because the kernel caches the
most recently used \texttt{vm\_area\_struct}, avoiding a search in the
red-black tree. In the continuous benchmark, this cached struct is always
conveniently used.

@@

The frag benchmark takes the most time of any benchmark. When a region of the
address space is being expanded, the kernel is forced to unmap it and map a new
area elsewhere in the address space. This operation needs creation of a new
\texttt{vm\_area\_struct}.  The kernel also must search through the red-black
tree, though it never finds a node which can be expanded to include the new
mapping. Then, an insertion on the tree is needed, possibly including
rebalancing. Moreover, the existing \texttt{ vm\_area\_struct} might need to be
deleted and removed from the tree. All of these operations ensure that frag
mmap pattern takes the most amount of time.

The strided workload is very similar to frag because it will also force the
kernel to always create a new structure and insert it into the tree after
traversing the tree first. However, it does not need to delete an existing
structure, because of which, it is faster than the frag benchmark.

Finally, the rand workload will also search through the red-black tree first,
but there is a reasonably high probability that it will find an existing
mapping adjacent to the desired mapping which can be expanded, obviating the
need to create a new struct and insert it into red-black tree. Therefore, most
of the times the time taken by rand benchmark is lower than strided.

@@

The overall behavior is as expected. In the random benchmark, as before, all
memory is exhausted quickly, and the benchmark induces swapping soon after the
benchmark begins. Many memory accesses thus take milliseconds to complete
because they require paging in or out, which is slow due to disk access.
Even so, the majority of benchmark run time comes from a minority of accesses.
Figure \ref{fig:swap_time_rand_cdf} shows that as much as 95\% of memory
accesses occur on the micro-second scale. Of the remaining 5\%, almost all
accesses complete in less than 20ms. Moreover, notice that the size of the
swap file makes very little difference in this workload because the disk has to
do random seeks regardless of swapfile size.

@@

We also find that in the continuous benchmark, a larger swap file actually has
slightly more of these slow memory accesses. We believe this has to do with the
fact that the large swapfile has less disk locality, and so causes more disk
seeks when paging out. Figure \ref{fig:swap_time_cdf} shows that with a larger
swap file, the continuous benchmark has a significantly higher number of
high-latency accesses as compared with the smaller swap file.

@@

Large swap files should be created with disk locality or on SSDs.

Looking at 99.99-th percentile latencies is important because they can have a
large impact on overall run time.
